Predicting Diamond Pricing: A Generalized Linear Model Analysis

### 2024-12-14 paper/paper.qmd
update the captions of graphs

We use the statistical programming language R [@citeR], alongside several key libraries and datasets to analyze and visualize data effectively. Our data [@diamonds_data] is sourced from the Data and Story Library, providing detailed attributes of diamonds for multiple regression analysis. Following @tellingstories, we consider how storytelling principles can enhance data interpretation and presentation. The tidyverse ecosystem [@citetidyverse] underpins our data manipulation and visualization, with dplyr [@citedplyr] for data wrangling and ggplot2 integrated packages to create visualizations. The here package [@citehere] simplifies file path management, while knitr [@citeknitr] and patchwork [@patchwork] streamline report generation and composite plotting. Bayesian modeling is conducted with rstanarm [@rstanarm], leveraging robust statistical frameworks. Additionally, caTools [@caTools] aids in statistical operations, and ggcorrplot [@ggcorrplot] facilitates correlation visualization.

The reason that I chose DASL is its directness and openness of data. The data set focuses on the diamond market, contains detailed 4C attributes and prices, directly supports the research objectives, and is completely open, ensuring the repeatability and transparency of the analysis. Additionally, there does not exist other data set about diamond price in the data set licensed in @tellingstories. However, we can find data about diamond prices from diamond retailers (such as Blue Nile or James Allen). However, these data sets usually come from commercial use, so they may have deviations from the inventory or pricing strategies of specific retailers. This deviation will limit the universality of the research results in a wider market. In addition, there may be some access restrictions to obtain the complete version of these data.

@@

The diamond price is a continuous numerical variable, ranging from close to 1000 USD to 10,000 USD. Through the histogram of @fig-price, we can find that the price of diamonds has obvious right skewness, and with the increase of price, the number of diamonds decreases gradually. In addition, combining the histogram and box chart in @fig-price, we can find that there is no obvious abnormal value of diamond prices. Most diamond prices are concentrated in the range of 2,500 USD to 5,000 USD, with an average price of about 4,000 USD and a median of 3,500 USD, indicating that a few high-priced diamonds have raised the overall average.

@@

The barplot on the left of @fig-color shows the distribution of diamond color from D to K. The barplot shows that the number of diamonds with color grades E and F is the largest, indicating that these nearly colorless diamonds are more popular in the market, probably because they have high quality and cost performance. In contrast, the number of completely colorless D-class diamonds and K-class diamonds with slight yellow is small, which may be due to their rarity and color deviation, respectively, resulting in low demand. This picture directly reflects the supply of diamonds of different color grades in the market and their potential demand trends.
The violin chart on the right of @fig-color shows the distribution of diamond prices in different color grades. The violio chart shows that the prices of diamonds with color grades D to F are generally higher, especially those with color grades D and E, which shows a wider price distribution range, indicating that the prices of high-end diamonds with these color grades can increase significantly. In contrast, the price distribution of diamonds with color grades I to K is more concentrated and the overall price is lower, reflecting that the price fluctuation of diamonds close to yellow is less in the market. In addition, there are significant outliers in top colors such as D and E, which may be due to the unusually high price of diamonds with large carats or other high-quality characteristics. This figure clearly shows the influence of color grade on price and its market value distribution characteristics.

@@

color_histogram_plot <- ggplot(cleaned_data, aes(x = color)) +
  geom_bar(fill = "pink", color = "black", alpha = 0.7) + 
  labs(
    title ="Distribution of Variable: Color",
    x = "Diamond Color",
    y = "Count") +
  theme_minimal() +
  theme(plot.title = element_text(size = 12, face = "bold",hjust = 0.5 ))

violin_plot_color <- ggplot(cleaned_data, aes(x = color, y = price)) +
  geom_violin(trim = FALSE, fill = "lightblue", alpha = 0.7) + 
  geom_boxplot(width = 0.2, outlier.color = "salmon", outlier.shape = 8, alpha = 0.7) + 
  labs(
    title = "Violin Plot of Price by Color",
    x = "Diamond Color",
    y = "Diamond Price (USD)"
  ) +
  theme_minimal()+
  theme(plot.title = element_text(size = 12, face = "bold",hjust = 0.5 ))

@@

The diamond carat size, a continuous numerical variable representing the weight of the diamond, ranges from approximately 0.30 to 2.02 carats in the dataset. One carat is equivalent to 200 milligrams, making it a key factor in determining the diamond's size and price.
The histogram on the left of @fig-carat_size shows the distribution of the Carat Size of diamonds. On the whole, it presents a right-skewed distribution, indicating that the number of small carats (such as 0.5 to 1.0 carats) is the largest in the market, while the number of larger carats is gradually decreasing. The red dotted line in the figure indicates the average carat size, and it can be seen that the carat size of most diamonds is concentrated on the left side of the average, which further reflects the dominant position of small carat diamonds in the market. This reflects that consumers have higher demand for small carats, while large carats are scarce because of their rarity and high price.
The scatter chart on the right of @fig-carat_size shows the relationship between the carat size of diamonds and the price. It can be clearly seen that the price rises quickly with the increase of carat size, showing a nonlinear growth trend. Especially in the range of close to 1.0 carats and larger carats, the price increase is more significant. This shows that the carat size has an important influence on the diamond price, but when the Dancla size exceeds a certain value (such as 1.5 carats), the price growth rate tends to be flat. In addition, the fitted black curve further clearly shows this nonlinear relationship, indicating that the marginal effect of carat size on price decreases.

@@

# Graph histogram
# Calculate the mean of diamond prices
mean_size <- mean(cleaned_data$carat_size, na.rm = TRUE)
carat_size_histogram_plot <- ggplot(cleaned_data, aes(x = carat_size)) +
  geom_histogram(fill = "palevioletred", color = "black", alpha = 0.7) +
  geom_vline(xintercept = mean_size, linetype = "dashed", color = "salmon") +
  labs(
    title = "Distribution of Variable: Carat Size",
    x = "Diamond Carat Size",
    y = "Count"
  ) +
  theme_minimal()+
  theme(plot.title = element_text(size = 12, face = "bold",hjust = 0.5 ))

plot_carat_size <- ggplot(cleaned_data, aes(x = carat_size, y = price)) +
  geom_point(color = "lightblue", alpha = 0.7) +
  geom_smooth(method = "loess", color = "black", se = FALSE) +
  labs(
    title = "Carat Size vs Price",
    x = "Carat Size",
    y = "Diamond Price (USD)"
  ) +
  theme_minimal()+
  theme(plot.title = element_text(size = 12, face = "bold",hjust = 0.5))

@@

The barplot on the left of @fig-clarity shows the distribution of diamond Clarity. It shows that the number of diamonds with clarity grades SI1 and SI2 is the largest, indicating that these grades of diamonds are the most widely available in the market, probably because they have higher cost performance and greater market demand. In contrast, the number of diamonds with clarity grade of IF(Internally Flawless) is the least, which reflects the rarity of completely flawless diamonds. This shows that diamonds with lower definition grades (such as SI1 and SI2) are the main ones in the market, while high definition diamonds are scarce.
The violin chart on the right of @fig-clarity shows the distribution of diamond prices at different clarity levels. With the improvement of clarity, the price rises significantly, especially for diamonds with high definition such as IF and VVS1. The price distribution is wider and there are obvious outliers, which may be caused by the carat size or other quality characteristics of these diamonds. On the other hand, the price distribution of grades with low definition (such as SI1 and SI2) is relatively concentrated, and the median is obviously lower than that of high-grade diamonds, but their price range shows their stable demand and high cost performance in the market.

@@

# Graph histogram
# Create the histogram with mean line
clarity_histogram_plot <- ggplot(cleaned_data, aes(x = clarity)) +
  geom_bar(bins = 20, fill = "pink", color = "black", alpha = 0.7) +
  labs(
    title = "Distribution of Variable: Clarity",
    x = "Diamond Clarity",
    y = "Count"
  ) +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  theme(plot.title = element_text(size = 12, face = "bold",hjust = 0.5))

violin_plot_clarity <- ggplot(cleaned_data, aes(x = clarity, y = price)) +
  geom_violin(trim = FALSE, fill = "lightblue", alpha = 0.7) + 
  geom_boxplot(width = 0.2, outlier.color = "salmon", outlier.shape = 8, alpha = 0.7) +
  labs(
    title = "Violin Plot of Price by Clarity",
    x = "Clarity",
    y = "Diamond Price(USD)"
  ) +
  theme_minimal()+
  theme(plot.title = element_text(size = 12, face = "bold",hjust = 0.5))

@@

The barplot on the left of @fig-cut shows the distribution of diamonds with different cutting grades. It indicates that the number of diamonds with Excellent and Very Good grades is obviously more, indicating that most diamonds on the market are concentrated in these high-cut grades, which may be because they have better visual effects and higher market demand. In contrast, the number of Good and Ideal diamonds is small, especially the rarity of Ideal cutting may reflect strict cutting ratio requirements and high quality standards.
The violin chart on the right of @fig-cut shows the price distribution of diamonds with different cutting grades. It shows the price range of each cutting grade is very close to the median, which shows that the cutting grade has little direct influence on the price. However, Ideal and Excellent diamonds are more obviously distributed in the high-end price range, which may be because these two grades of diamonds are usually combined with other high-quality characteristics, such as high carat number or clarity. At the same time, the overall price range is relatively large, indicating that other factors (such as carats, colors, etc.) may play a more important role in determining the price.

@@

# Graph histogram
# Create the histogram with mean line
cut_histogram_plot <- ggplot(cleaned_data, aes(x = cut)) +
  geom_bar(bins = 20, fill = "palevioletred", color = "black", alpha = 0.7)+
  labs(
    title = "Distribution of Variable: Cut",
    x = "Diamond Cut",
    y = "Count"
  ) +
  theme_minimal()+
  theme(plot.title = element_text(size = 12, face = "bold",hjust = 0.5))

violin_plot_cut <- ggplot(cleaned_data, aes(x = cut, y = price)) +
  geom_violin(trim = FALSE, fill = "lightblue", alpha = 0.7) + 
  geom_boxplot(width = 0.2, outlier.color = "salmon", outlier.shape = 8, alpha = 0.7) +
  labs(
    title = "Violin Plot of Price by Cut",
    x = "Diamond Cut",
    y = "Diamond Price (USD)"
  ) +
  theme_minimal()+
  theme(plot.title = element_text(size = 12, face = "bold",hjust = 0.5))

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.pos="H"}
#| label: fig-correlation_heat
#| fig-cap: "Correlation: Heat Graph"

@@

Despite its advantages, the GLM has potential limitations and may not be suitable in certain scenarios. When the distribution of response variable is hard to observe, we may not able to assume its distribution. Furthermore, highly nonlinear relationships may not be adequately captured by the specified link function, requiring alternative modeling approaches.

@@

The GLM model captures the general trend of the transformed prices well, especially in the lower and middle ranges. The predicted values closely follow the transformed prices in these areas, indicating that the GLM model performs well for most of the data. However, at higher transformed price levels, the predictions tend to underestimate the values, as indicated by the wider spread of red points compared to the smoother blue line. The variability of the transformed prices increases at higher indices, but the model does not fully capture this. This suggests that the GLM model, with its fixed Gamma distribution and log link function, may oversimplify the behavior of the response variable in these regions. The underestimation at higher levels highlights potential limitations in the model’s assumptions.

>>>>

We use the statistical programming language R [@citeR], alongside several key libraries and data sets to analyze and visualize data effectively. Our data [@diamonds_data] is sourced from the Data and Story Library, providing detailed attributes of diamonds for multiple regression analysis. Following @tellingstories, we consider how storytelling principles can enhance data interpretation and presentation. The tidyverse ecosystem [@citetidyverse] underpins our data manipulation and visualization, with dplyr [@citedplyr] for data wrangling and ggplot2 integrated packages to create visualizations. The here package [@citehere] simplifies file path management, while knitr [@citeknitr] and patchwork [@patchwork] streamline report generation and composite plotting. Bayesian modeling is conducted with rstanarm [@rstanarm], leveraging robust statistical frameworks. Additionally, caTools [@caTools] aids in statistical operations, and ggcorrplot [@ggcorrplot] facilitates correlation visualization.

The reason that I chose DASL is its directness and openness of data. The data set focuses on the diamond market, contains detailed 4C attributes and prices, directly supports the research objectives, and is completely open, ensuring the reproducibility and transparency of the analysis. Additionally, there does not exist other data set about diamond price in the data set licensed in @tellingstories. However, we can find data about diamond prices from diamond retailers (such as Blue Nile or James Allen). However, these data sets usually come from commercial use, so they may have deviations from the inventory or pricing strategies of specific retailers. This deviation will limit the universality of the research results in a wider market. In addition, there may be some access restrictions to obtain the complete version of these data.

@@

The diamond price is a continuous numerical variable, ranging from close to 1000 USD to 10,000 USD. Through the left graph of @fig-price, we can find that the price of diamonds has obvious right skewness, and with the increase of price, the number of diamonds decreases gradually. In addition, in @fig-price, we can find that there is no obvious abnormal value of diamond prices. Most diamond prices are concentrated in the range of 2,500 USD to 5,000 USD, with an average price of about 4,000 USD and a median of 3,500 USD, indicating that a few high-priced diamonds have raised the overall average.

@@

The left graph of @fig-color shows the distribution of diamond color from D to K. It shows that the number of diamonds with color grades E and F is the largest, indicating that these nearly colorless diamonds are more popular in the market, probably because they have high quality and cost performance. In contrast, the number of completely colorless D-class diamonds and K-class diamonds with slight yellow is small, which may be due to their rarity and color deviation, respectively, resulting in low demand. This picture directly reflects the supply of diamonds of different color grades in the market and their potential demand trends.
The right graph of @fig-color shows the distribution of diamond prices in different color grades. It shows that the prices of diamonds with color grades D to F are generally higher, especially those with color grades D and E, which shows a wider price distribution range, indicating that the prices of high-end diamonds with these color grades can increase significantly. In contrast, the price distribution of diamonds with color grades I to K is more concentrated and the overall price is lower, reflecting that the price fluctuation of diamonds close to yellow is less in the market. In addition, there are significant outliers in top colors such as D and E, which may be due to the unusually high price of diamonds with large carats or other high-quality characteristics. It clearly shows the influence of color grade on price and its market value distribution characteristics.

@@

color_histogram_plot <- ggplot(cleaned_data, aes(x = color)) +
  geom_bar(fill = "pink", color = "black", alpha = 0.7) + 
  labs(
    title ="Distribution of Variable: Color",
    x = "Diamond Color",
    y = "Count") +
  theme_minimal() +
  theme(plot.title = element_text(size = 10, face = "bold",hjust = 0.5 ))

violin_plot_color <- ggplot(cleaned_data, aes(x = color, y = price)) +
  geom_violin(trim = FALSE, fill = "lightblue", alpha = 0.7) + 
  geom_boxplot(width = 0.2, outlier.color = "salmon", outlier.shape = 8, alpha = 0.7) +
  labs(
    title = "Price Distribution\nAcross Diamond Colors",
    x = "Diamond Color",
    y = "Diamond Price (USD)"
  ) +
  theme_minimal()+
  theme(plot.title = element_text(size = 10, face = "bold",hjust = 0.5 ))

@@

The diamond carat size, a continuous numerical variable representing the weight of the diamond, ranges from approximately 0.30 to 2.02 carats in the data set. One carat is equivalent to 200 milligrams, making it a key factor in determining the diamond's size and price.
The left graph of @fig-carat_size shows the distribution of the Carat Size of diamonds. On the whole, it presents a right-skewed distribution, indicating that the number of small carats (such as 0.5 to 1.0 carats) is the largest in the market, while the number of larger carats is gradually decreasing. The red dotted line in the figure indicates the average carat size, and it can be seen that the carat size of most diamonds is concentrated on the left side of the average, which further reflects the dominant position of small carat diamonds in the market. This reflects that consumers have higher demand for small carats, while large carats are scarce because of their rarity and high price.
The right graph of @fig-carat_size shows the relationship between the carat size of diamonds and the price. It can be clearly seen that the price rises quickly with the increase of carat size, showing a nonlinear growth trend. Especially in the range of close to 1.0 carats and larger carats, the price increase is more significant. This shows that the carat size has an important influence on the diamond price, but when the Diamond size exceeds a certain value (such as 1.5 carats), the price growth rate tends to be flat. In addition, the fitted black curve further clearly shows this nonlinear relationship, indicating that the marginal effect of carat size on price decreases.

@@

# Graph histogram
# Calculate the mean of diamond prices
mean_size <- mean(cleaned_data$carat_size, na.rm = TRUE)
carat_size_histogram_plot <- ggplot(cleaned_data, aes(x = carat_size)) +
  geom_histogram(fill = "palevioletred", color = "black", alpha = 0.7) +
  geom_vline(xintercept = mean_size, linetype = "dashed", color = "salmon") +
  labs(
    title = "Distribution of Variable: Carat Size",
    x = "Diamond Carat Size",
    y = "Count"
  ) +
  theme_minimal()+
  theme(plot.title = element_text(size = 10, face = "bold",hjust = 0.5 ))

plot_carat_size <- ggplot(cleaned_data, aes(x = carat_size, y = price)) +
  geom_point(color = "lightblue", alpha = 0.7) +
  geom_smooth(method = "loess", color = "black", se = FALSE) +
  labs(
    title = "Carat Size vs Price",
    x = "Carat Size",
    y = "Diamond Price (USD)"
  ) +
  theme_minimal()+
  theme(plot.title = element_text(size = 10, face = "bold",hjust = 0.5))

@@

The left graph of @fig-clarity shows the distribution of diamond Clarity. It shows that the number of diamonds with clarity grades SI1 and SI2 is the largest, indicating that these grades of diamonds are the most widely available in the market, probably because they have higher cost performance and greater market demand. In contrast, the number of diamonds with clarity grade of IF(Internally Flawless) is the least, which reflects the rarity of completely flawless diamonds. This shows that diamonds with lower definition grades (such as SI1 and SI2) are the main ones in the market, while high definition diamonds are scarce.
The right graph of @fig-clarity shows the distribution of diamond prices at different clarity levels. With the improvement of clarity, the price rises significantly, especially for diamonds with high definition such as IF and VVS1. The price distribution is wider and there are obvious outliers, which may be caused by the carat size or other quality characteristics of these diamonds. On the other hand, the price distribution of grades with low definition (such as SI1 and SI2) is relatively concentrated, and the median is obviously lower than that of high-grade diamonds, but their price range shows their stable demand and high cost performance in the market.

@@

# Graph histogram
# Create the histogram with mean line
clarity_histogram_plot <- ggplot(cleaned_data, aes(x = clarity)) +
  geom_bar(bins = 20, fill = "pink", color = "black", alpha = 0.7) +
  labs(
    title = "Distribution of Variable: Clarity",
    x = "Diamond Clarity",
    y = "Count"
  ) +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  theme(plot.title = element_text(size = 10, face = "bold",hjust = 0.5))

violin_plot_clarity <- ggplot(cleaned_data, aes(x = clarity, y = price)) +
  geom_violin(trim = FALSE, fill = "lightblue", alpha = 0.7) + 
  geom_boxplot(width = 0.2, outlier.color = "salmon", outlier.shape = 8, alpha = 0.7) +
  labs(
    title = "Price Distribution\nAcross Diamond Clarity",
    x = "Clarity",
    y = "Diamond Price(USD)"
  ) +
  theme_minimal()+
  theme(plot.title = element_text(size = 10, face = "bold",hjust = 0.5))

@@

The left graph of @fig-cut shows the distribution of diamonds with different cutting grades. It indicates that the number of diamonds with Excellent and Very Good grades is obviously more, indicating that most diamonds on the market are concentrated in these high-cut grades, which may be because they have better visual effects and higher market demand. In contrast, the number of Good and Ideal diamonds is small, especially the rarity of Ideal cutting may reflect strict cutting ratio requirements and high quality standards.
The right graph of @fig-cut shows the price distribution of diamonds with different cutting grades. It shows the price range of each cutting grade is very close to the median, which shows that the cutting grade has little direct influence on the price. However, Ideal and Excellent diamonds are more obviously distributed in the high-end price range, which may be because these two grades of diamonds are usually combined with other high-quality characteristics, such as high carat number or clarity. At the same time, the overall price range is relatively large, indicating that other factors (such as carats, colors, etc.) may play a more important role in determining the price.

@@

# Graph histogram
# Create the histogram with mean line
cut_histogram_plot <- ggplot(cleaned_data, aes(x = cut)) +
  geom_bar(bins = 20, fill = "palevioletred", color = "black", alpha = 0.7)+
  labs(
    title = "Distribution of Variable: Cut",
    x = "Diamond Cut",
    y = "Count"
  ) +
  theme_minimal()+
  theme(plot.title = element_text(size = 10, face = "bold",hjust = 0.5))

violin_plot_cut <- ggplot(cleaned_data, aes(x = cut, y = price)) +
  geom_violin(trim = FALSE, fill = "lightblue", alpha = 0.7) + 
  geom_boxplot(width = 0.2, outlier.color = "salmon", outlier.shape = 8, alpha = 0.7) +
  labs(
    title = "Price Distribution\nAcross Diamond Cut",
    x = "Diamond Cut",
    y = "Diamond Price (USD)"
  ) +
  theme_minimal()+
  theme(plot.title = element_text(size = 10, face = "bold",hjust = 0.5))

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.pos="H"}
#| label: fig-correlation_heat
#| fig-cap: "Correlation Matrix of Diamond Attributes"

@@

While GLM has advantages, it may face limitations when response variable distributions are unclear. Furthermore, highly nonlinear relationships may not be adequately captured by the specified link function, requiring alternative modeling approaches.

@@

The GLM model captures the general trend of the transformed prices well, especially in the lower and middle ranges. The predicted values closely follow the transformed prices in these areas, indicating that the GLM model performs well for most of the data. However, at higher transformed price levels, the predictions tend to underestimate the values, as indicated by the wider spread of red points compared to the smoother blue line. The variability of the transformed prices increases at higher indices, but the model does not fully capture this. This suggests that the GLM model’s fixed Gamma distribution and log link may oversimplify response variable behavior, leading to underestimation in higher price ranges.
